---
title: 'NOAA Storm Database analysis: Coursera Reproducible Research Project 2'
author: "Gustavo Recio"
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

# Synopsis
Severe weather conditions are known to be the cause of public concern and economic consequences.

The aim of this project is to explore the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database, which tracks several characteristics of weather events in the United States, including their location, date, and their consequences, such as personal injuries and property damage.

As part of this report we will address the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

After aggregating the data by type of event, we can see the following outcomes:

1. For the public health (both for injuries and fatalities), tornados cause the biggest damage
2. In terms of economic consequences, floods cause the biggest repercussions


# Data processing
Downloading the data source:
```{r data_download}
destFile <- "storm-data.csv.bz2" 
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"   
if (!file.exists(destFile)) {
    download.file(fileURL, destFile, method="auto")
}
```

Loading the file into a data frame
```{r data_load, cache=TRUE}
sourceData <- read.csv(bzfile(destFile), sep=",", header = TRUE)
```

Initial analysis of the data
```{r data_preview, cache=TRUE}
names(sourceData)
head(sourceData)
```

Not all the columns in the dataframe are relevant to answer the current query. Removing unused columns
```{r remove_columns, cache=TRUE}
interestingVariables = c('EVTYPE', 'FATALITIES', 'INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP')
subsetData <- select(sourceData, interestingVariables)
names(subsetData)
```

Entries in the variables PROPDMG and CROPDMG are not directly processable, as they have different units. They need to be preprocessed
```{r preprocess_df, cache=TRUE}
preprocessedData <- subsetData %>%
  mutate(PROPDMG_SAME_UNIT =
            if_else(PROPDMGEXP == 'H', PROPDMG * 10^2,
            if_else(PROPDMGEXP == 'K', PROPDMG * 10^3,
            if_else(PROPDMGEXP == 'M', PROPDMG * 10^6,
            if_else(PROPDMGEXP == 'B', PROPDMG * 10^9,
            0))))) %>%
  mutate(CROPDMG_SAME_UNIT =
            if_else(CROPDMGEXP == 'H', CROPDMG * 10^2,
            if_else(CROPDMGEXP == 'K', CROPDMG * 10^3,
            if_else(CROPDMGEXP == 'M', CROPDMG * 10^6,
            if_else(CROPDMGEXP == 'B', CROPDMG * 10^9,
            0)))))
```
# Results
## Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r create_personal_injury_graphic}
fatalities <- preprocessedData %>%
  group_by(EVTYPE) %>%
  summarize(TOTAL_FATALITIES = sum(FATALITIES, na.rm = TRUE)) %>%
  arrange(desc(TOTAL_FATALITIES)) %>%
  filter(row_number() <= 10)

fatalitiesPlot <- ggplot(fatalities, aes(x = EVTYPE, y = TOTAL_FATALITIES)) + 
  geom_bar(stat = "identity", fill = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("fatalities") +
  ggtitle("Fatalities per event type")


injuries <- preprocessedData %>%
  group_by(EVTYPE) %>%
  summarize(TOTAL_INJURIES = sum(INJURIES, na.rm = TRUE)) %>%
  arrange(desc(TOTAL_INJURIES)) %>%
  filter(row_number() <= 10)

injuriesPlot <- ggplot(injuries, aes(x = EVTYPE, y = TOTAL_INJURIES)) + 
  geom_bar(stat = "identity", fill = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("injuries") +
  ggtitle("Injuries per event type")


# Multiplot function copied from: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
multiplot(fatalitiesPlot, injuriesPlot, cols=2)
```


## Across the United States, which types of events have the greatest economic consequences?

First, let's explore the property and crop damage separatedly
```{r create_economic_impact_graphic}
propertyDamage <- preprocessedData %>%
  group_by(EVTYPE) %>%
  summarize(TOTAL_PROP_DMG = sum(PROPDMG_SAME_UNIT, na.rm = TRUE)) %>%
  arrange(desc(TOTAL_PROP_DMG))
  

propertyDamagePlot <- ggplot(filter(propertyDamage, row_number() <= 10), aes(x = EVTYPE, y = TOTAL_PROP_DMG)) + 
  geom_bar(stat = "identity", fill = "blue") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("property damage") +
  ggtitle("Property damage per event type")


cropDamage <- preprocessedData %>%
  group_by(EVTYPE) %>%
  summarize(TOTAL_CROP_DMG = sum(CROPDMG_SAME_UNIT, na.rm = TRUE)) %>%
  arrange(desc(TOTAL_CROP_DMG))

cropDamagePlot <- ggplot(filter(cropDamage, row_number() <= 10), aes(x = EVTYPE, y = TOTAL_CROP_DMG)) + 
  geom_bar(stat = "identity", fill = "blue") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("crop damage") +
  ggtitle("Crop damage per event type")


# Multiplot function copied from: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
multiplot(propertyDamagePlot, cropDamagePlot, cols=2)
```

In order to better answer the question, we have to combine both property and crop damage
```{r combined}
totalDamage <- inner_join(propertyDamage, cropDamage) %>%
  mutate(TOTAL_DAMAGE = TOTAL_PROP_DMG + TOTAL_CROP_DMG) %>%
  arrange(desc(TOTAL_PROP_DMG)) %>%
  filter(row_number() <= 10)

ggplot(filter(totalDamage, row_number() <= 10), aes(x = EVTYPE, y = TOTAL_CROP_DMG)) + 
  geom_bar(stat = "identity", fill = "blue") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Total damage") +
  ggtitle("Total damage")
```